Open archaeology, open source?

### 2024-02-06 analysis/_01-introduction.qmd
Minor edits to readability and text flow

<!--
* There is open archaeology
* Open archaeology is related to open source
* Is open archaeology open source, and is open source open?
  * [Is 'open archaeology' just uploading CSVs?]
-->

@@

However, academic open source has a complicated relationship with open source as practiced by professional software developers, which has its own distinct history and is framed by different objectives, challenges, and value regimes. Despite this, the open science movement, within which open archaeology emerged, draws direct inspiration from open source. For instance, the Open Knowledge Foundation [-@openknowledgefoundation2015] publishes a widely accepted definition of "open" in the context of scholarly communication that explicitly refers to the definition of "open source" published by the Open Source Initiative [-@opensourceinitiative2007], an authoritative open source advocacy group. The open science movement further mimics open source by operationalizing scholarly communication through technical infrastructures and protocols that closely resemble systems and processes designed to develop open source software (e.g., the use of plain text, line-resolution version control, emphasis on formal licencing, the general hacker aesthetic). However, academic work, including the development of academic software, differs significantly from the work involved in massive open source projects that literally run the internet, such as the Linux kernel, openSSL and the Firefox web browser. While they may use similar tools and technical protocols to manage coding operations, the open science and open source movements are governed by different social and professional warrants and interests. In other words, publishing code openly on the web has different meanings, impacts and implications for archaeologists and professional software developers [@ratto2007; @kelty2008 chapter 9].

The relationship between open science and open source is also complicated by rhetorical claims that have a questionable connection to how academics actually _do_ open source. 
Does academic open source actually make research processes more transparent and improve research outcomes? 
Is it actually boosting efficiency by establishing a common store of knowledge and productive code? 
Is it actually helping to foster new globe-spanning connections and lead to novel research trajectories that would not otherwise come to pass?
Basically, is it more than just uploading code and data files to the internet?

@@

That being said, archaeologists are prolific software developers. This is made evident by open-archaeo, a comprehensive list of open source archaeological software and resources that indexes 495 records, as of the time of this article's publication. But beyond simply making their code available on the web, do archaeologists also implement social strategies to advance open source ideals? Does archaeological open source actually help achieve greater transparency, sustainability, and community participation? And if not, what does it actually achieve?

@@

Our statistical and network analyses consider how archaeological software development may be benefiting from or missing out on the affordances that open source development models provide, specifically the value added through working as part of a broader community of invested stakeholders, processes of iterative improvement, and increased code transparency. We found that the vast majority of projects lack any collective effort and activity tends to abruptly end shortly after work is initiated. Projects that are maintained by multiple contributors tend to be backed by funded initiatives who employ locally and socially engaged individuals, or are warranted by participation in a scholarly community exhibiting genuine need for a particular kind of resource. Moreover, we found that the individuals who play critical roles in supporting the archaeological open source community are precariously employed workers. Contrary to popular claims about open source being inherently distributed, resilient, and open-ended, we found that, overall, archaeological open source is actually quite centralized, fragile, and based nearly exclusively on existing professional connections and endeavours.

>>>>

@@

The relationship between open science and open source is complicated by rhetorical claims that have a questionable connection to how academics actually _do_ open source. 
Does academic open source actually make research processes more transparent and improve research outcomes? 
Is it actually boosting efficiency by establishing a common store of knowledge and productive code? 
Is it actually helping to foster new globe-spanning connections and lead to novel research trajectories that would not otherwise come to pass?
Basically, is there more to it than just uploading code and data files to the internet?

@@

There is no question that archaeologists are prolific software developers [@batist2023].
But beyond simply making their code available on the web, do archaeologists also implement social strategies to advance open source ideals? Does archaeological open source actually help achieve greater transparency, sustainability, and community participation? And if not, what does it actually achieve?

@@

We use quantitative analysis to consider how archaeological software development may be benefiting from, or missing out on, the affordances that open source development models provide, specifically the value added through working as part of a broader community of invested stakeholders, processes of iterative improvement, and increased code transparency.

### 2024-02-06 analysis/_02-background.qmd
Minor edits to readability and text flow

<!--
* What is open source?
* Summary of key publications in open archaeology / DIY archaeology
* But is any of this real? There seems to be a lack of reflection about what open archaeology is.
  -->

@@

Open science is typically compared with the open source movement in that they both involve a distributed, digitally-mediated and worldwide labour force, who somehow derive rough consensus directed towards assets held in the public domain [@tennant2020]. But they differ in terms of the contexts in which they operate, the stakeholders involved, and the kinds of outcomes they produce. Whereas open source emerged from concern of consumer rights and then developed as a means of maintaining resilient and collectively motivated projects, open science comes out of a desire to make research practices more transparent and accessible. Open source is performed by professional and hobbyist software developers alike, and participants contribute in a wide variety of ways (including: programming, writing documentation, translating software and documentation, bug reporting, and financial support), but, in open science, scientists are usually the only participants actively involved in creating and maintaining contributions. Moreover, whereas open source projects often attract participants with varied stakes in the software and use cases in mind, open science projects are typically bounded by small communities of specialists with very particular needs, and rarely anticipate use of their work beyond the limited scope of their primary targeted audience. Additionally, open science is bounded by the professional contexts in which science operates, and as such, produces outputs that can be easily credited to specific sets of individuals for reasons of tenure and promotion. Open science projects whose contributions are supported by research funding also face sustainability concerns, as participants lose motivation to contribute once funding runs out. Once a project is completed, papers have been published, and credit has been allocated, it is common for scientists to mark their projects as finished and move on to new endeavours. Open source projects, on the other hand, are motivated by a more practical need for the software to function properly in perpetuity, and contributors may remain actively or sporadically involved to satisfy users' needs, or to direct users to derivative and functional forks of abandoned software.

@@

## Open Source Infrastructure

Open source is an inherently internet-based development model and is supported by technical infrastructures which facilitate global distribution of labour and code.
Here we provide a brief overview of key technologies that archaeologists have come to rely on as they develop open source software.

@@

Forges, or collaborative software development platforms like GitHub, GitLab and Codeberg are designed to facilitate open source software development by hosting public git repositories.
However, they also support common software developer and project management practices, such as issue and bug tracking, code-commenting, task management, identity and permissions management, web publishing, vulnerability detection, creation and maintenance of metadata, and financial sponsorship.
These platforms also implement standard social media functions, like the ability to follow projects and individual users to receive updates on their activities, "star" certain repositories as favourites, and maintain a public-facing profile that includes personally-identifying information (e.g. profile picture, username, real name, employer or affiliation), references to all public activity on the platform, and links to the user's other social media profiles.
Code-sharing platforms thus serve as comprehensive developer portfolios and community networking resources.
While these additional features are meant to complement and enhance the experience of contributing to open source projects, they are not actually part of the git protocol.
See @tbl-glossary for a glossary of terminology used to describe many of these concepts and features, and which we use throughout this paper.

>>>>

Academic open source has a complicated relationship with open source as practiced by professional software developers, which has its own distinct history and is framed by different objectives, challenges, and value regimes. Despite this, the open science movement, within which open archaeology emerged, draws direct inspiration from open source. For instance, the Open Knowledge Foundation [-@openknowledgefoundation2015] publishes a widely accepted definition of "open" in the context of scholarly communication that explicitly refers to the definition of "open source" published by the Open Source Initiative [-@opensourceinitiative2007], an authoritative open source advocacy group. The open science movement further mimics open source by operationalizing scholarly communication through technical infrastructures and protocols that closely resemble systems and processes designed to develop open source software (e.g., the use of plain text, line-resolution version control, emphasis on formal licencing, the general hacker aesthetic). However, academic work, including the development of academic software, differs significantly from the work involved in massive open source projects that literally run the internet, such as the Linux kernel, openSSL and the Firefox web browser. While they may use similar tools and technical protocols to manage coding operations, the open science and open source movements are governed by different social and professional warrants and interests. In other words, publishing code openly on the web has different meanings, impacts and implications for archaeologists and professional software developers [@ratto2007; @kelty2008 chapter 9].

@@

Open science is typically compared with the open source movement in that they both involve a distributed, digitally-mediated and worldwide labour force, who somehow derive rough consensus directed towards assets held in the public domain [@tennant2020]. But they differ in terms of the contexts in which they operate, the stakeholders involved, and the kinds of outcomes they produce. Whereas open source emerged from concern of consumer rights and then developed as a means of maintaining resilient and collectively motivated projects, open science comes out of a desire to make research practices more transparent and accessible. Open source is performed by professional and hobbyist software developers alike, and participants contribute in a wide variety of ways (including: programming, writing documentation, translating software and documentation, bug reporting, and financial support), but, in open science, scientists are usually the only participants actively involved in creating and maintaining contributions. 
<!-- TODO: Some strong claims here. Can we add citations? -->Moreover, whereas open source projects often attract participants with varied stakes in the software and use cases in mind, open science projects are typically bounded by small communities of specialists with very particular needs, and rarely anticipate use of their work beyond the limited scope of their primary targeted audience. Additionally, open science is bounded by the professional contexts in which science operates, and as such, produces outputs that can be easily credited to specific sets of individuals for reasons of tenure and promotion. Open science projects whose contributions are supported by research funding also face sustainability concerns, as participants lose motivation to contribute once funding runs out. Once a project is completed, papers have been published, and credit has been allocated, it is common for scientists to mark their projects as finished and move on to new endeavours. Open source projects, on the other hand, are motivated by a more practical need for the software to function properly in perpetuity, and contributors may remain actively or sporadically involved to satisfy users' needs, or to direct users to derivative and functional forks of abandoned software.

@@

## Git and GitHub

Open source is an inherently internet-based development model and is supported by technical infrastructures which facilitate global distribution of labour and code.
Here we provide a brief overview of key technologies that archaeologists have come to rely on as they develop open source software.
See @tbl-glossary a glossary of the git-, GitHub- and software engineering-related terminology which we use here and throughout this paper.

@@

Software forges—collaborative web platforms like GitHub, GitLab and Codeberg—are designed to facilitate open source software development by hosting public git repositories.
However, they also support common software developer and project management practices, such as issue and bug tracking, code-commenting, task management, identity and permissions management, web publishing, vulnerability detection, creation and maintenance of metadata, and financial sponsorship.
These platforms also implement standard social media functions, like the ability to follow projects and individual users to receive updates on their activities, "star" certain repositories as a combine bookmarking and 'like' feature, and maintain a public-facing profile that includes personally-identifying information (e.g. profile picture, username, real name, employer or affiliation), references to all public activity on the platform, and links to the user's other social media profiles.
Code-sharing platforms thus serve as comprehensive developer portfolios and community networking resources.
While these additional features are meant to complement and enhance the experience of contributing to open source projects, they are not actually part of the git protocol.

### 2024-02-06 analysis/_03-data_methodology.qmd
Minor edits to readability and text flow

```{r data-oarch}
# open-archaeo version 2024-02-03
# https://doi.org/10.5281/zenodo.10613694
oarch <- read_csv(here("analysis/data/raw_data/open-archaeo_2024-02-04.csv"))

>>>>

```{r data-oarch}
# open-archaeo version 2024-02-06
# https://doi.org/10.5281/zenodo.10625236
oarch <- read_csv(here("analysis/data/raw_data/open-archaeo_2024-02-06.csv"))

### 2024-02-06 analysis/_04-open_archaeology.qmd
Revised with 2024-02-04 version.
Updated the section on average centrality by platform, and another very minor reference correction.
Minor edits to readability and text flow

<!-- Results I: description of open archaeology projects

- What kind of projects comprise open archaeology?
  x Breakdown of tags and categories
  x Breakdown of platforms and languages
  x Tags by category
  x Licenses?
- How has this changed over time?
  - Growth of open archaeology by number of commits
  - Growth of open archaeology by number of repos
  - Compare changing proportions of each category on a year by year basis
  - Identify periods associated with different rates of growth, and relate these with different general attitudes in the history of digital archaeology
-->

As of writing, open-archaeo catalogues `r nrow(oarch)` resources created by and for archaeologists.
It includes both software and documents, but not research compendiums.^[see [https://github.com/benmarwick/ctv-archaeology#publications-that-include-r-code](https://github.com/benmarwick/ctv-archaeology#publications-that-include-r-code) for a similar list of archaeology publications that include R code]
We annotated each record with categories and tags that describe what aspect of archaeological research each tool or resource was meant to address.
We categorized all records based on how each tool or resource is meant to be accessed or used, and assigned tags based on how developers identified their projects' purpose and scope.
See [@batist2023] for a more comprehensive overview of the tags and categories applied to open-archaeo.

@@

In our breakdown of open-archaeo by category (see @tbl-categories), we demonstrate the pervasiveness of various development models, and the requisite technical capabilities that developers assume users hold.
Most resources (`r percent(p_platform)`) included in open-archaeo are designed to be used atop an existing "platform" -- for example a package that extends a programming language or a plugin for an application.
The designers of this code are basically creating additional functions within the base platform that are useful for archaeological purposes.
Others create standalone software (`r percent(p_standalone)`) that can be run independently of such platforms, for example desktop or web apps.
A significant number of projects also comprise of datasets (`r percent(p_datasets)`) and non-packaged code snippets (`r percent(p_scripts)`) that have been made available for general use.

@@

Our analysis of tags highlights aspects of archaeological work that software developers are inclined to contribute to (@tbl-tags).
The most common tags unsurprisingly deal with work that naturally benefits from advanced information processing afforded by computers, such as statistical analysis, sample calibration, geographical analysis, data management, and chronological modelling.
Educational resources and practical guides are also well represented due to the web's usefulness as a medium for sharing and communication.

When we compare categories with tags, we see the general domains that each kind of resource is designed to serve.
We see that packages are fairly common across the board.
Tags that are notable for having a higher proportion of standalone software include archaeogenetics, data management, 3D modelling, photogrammetry, drivers and IO, and simulations or agent based modelling.
These tools may require greater access to system resources, or may require more complex user interfaces than are more complex than what R or Python IDEs (integrated development environments) tend to provide.

@@

Roughly half of open-archaeo repositories are accompanied by an explicit license (@tbl-repo-licenses).
Two common free software licenses account for the majority of these: the GNU General Public License (GPL, `r percent(p_license_gpl)`) and the MIT License (`r percent(p_license_mit)`).
These differ primarily in the restrictions they place on reuse: the MIT License aims to be maximally permissive, while the GPL is a 'copyleft' license that specifies that all derivative works must be distributed under similar terms (in other words, it prohibits the use of open source software within non-open software, [@dusollier2007]).
Archaeologists' preference for the more restrictive of these two licenses is the reverse of the general trend seen in open source projects on GitHub [@balter2015].
Creative Commons licenses are a distant third place (`r percent(p_license_cc)` of repositories), in contrast to their widespread use for other forms of scholarly output [@kim2007].
Many repositories do not specify a license; given a documented misconception among academics that GitHub can serve as a sustainable and long-term code and data hosting platform [@milliken2021; @escamilla2022; @escamilla2023], it is possible that many maintainers whose work is included in open-archaeo similarly assumed that making their work available, without explicitly stating permissible use, is enough to allow unrestricted access to the repository's contents.
However, we can not verify this potential explanation given the methods we currently employ, and more discursive qualitative research is needed to explore the rationales behind such decisions.

>>>>

As of writing, open-archaeo catalogues `r nrow(oarch)` resources created by and for archaeologists.
It includes both software and documents, but not research compendiums.^[see [https://github.com/benmarwick/ctv-archaeology#publications-that-include-r-code](https://github.com/benmarwick/ctv-archaeology#publications-that-include-r-code) for a similar list of archaeology publications that include R code]
We annotated each record with categories and tags that describe what aspect of archaeological research each tool or resource was meant to address.
We also categorized all records based on how each tool or resource is meant to be accessed or used, and assigned tags based on how developers identified their projects' purpose and scope.
See @batist2023 for a more comprehensive overview of the tags and categories applied to open-archaeo.

@@

In our breakdown of open-archaeo by category (see @tbl-categories), we demonstrate the pervasiveness of various development models, and the requisite technical capabilities that developers assume users hold.
Most resources (`r percent(p_platform)`) included in open-archaeo are designed to be used atop an existing "platform" -- for example a package that extends a programming language or a plugin for an application.
Essentially such projects create additional functions within the base platform that are useful for archaeological purposes.
Others create standalone software (`r percent(p_standalone)`) that can be run independently of such platforms, for example desktop or web apps.
A significant number of projects also comprise of datasets (`r percent(p_datasets)`) and non-packaged code snippets (`r percent(p_scripts)`) that have been made available for general use.

@@

Our analysis of thematic tags highlights aspects of archaeological work that software developers are inclined to contribute to (@tbl-tags).
The most common themes are work that naturally benefits from advanced information processing afforded by computers, such as statistical analysis, sample calibration, geographical analysis, data management, and chronological modelling.
Educational resources and practical guides are also well represented due to the web's usefulness as a medium for sharing and communication.

When we compare categories with thematic tags, we see the general domains that each kind of resource is designed to serve.
We see that packages are fairly common across the board.
Tags that are notable for having a higher proportion of standalone software include archaeogenetics, data management, 3D modelling, photogrammetry, drivers and IO, and simulations or agent based modelling.
These tools may require greater access to system resources, or may require more complex user interfaces than are more complex than what R or Python IDEs (integrated development environments) tend to provide.

@@

Roughly half of open-archaeo repositories are accompanied by an explicit license (@tbl-repo-licenses).
Two common free software licenses account for the majority of these: the GNU General Public License (GPL, `r percent(p_license_gpl)`) and the MIT License (`r percent(p_license_mit)`).
These differ primarily in the restrictions they place on reuse: the MIT License aims to be maximally permissive, while the GPL is a 'copyleft' license that specifies that all derivative works must be distributed under similar terms (in other words, it prohibits the use of open source software within non-open software, [@dusollier2007].
Interestingly, archaeologists' preference for the more restrictive of these two licenses is the reverse of the general trend seen in open source projects on GitHub [@balter2015].
Creative Commons licenses are a distant third place (`r percent(p_license_cc)` of repositories), in contrast to their widespread use for other forms of scholarly output [@kim2007].
Many repositories do not specify a license; given a documented misconception among academics that GitHub can serve as a sustainable and long-term code and data hosting platform [@milliken2021; @escamilla2022; @escamilla2023], it is possible that many maintainers whose work is included in open-archaeo similarly assumed that making their work available, without explicitly stating permissible use, is enough to allow unrestricted access to the repository's contents.
However, we can not verify this potential explanation given the methods we currently employ, and more discursive qualitative research is needed to explore the rationales behind such decisions.

### 2024-02-06 analysis/_05-collaboration.qmd
General updates to the text, and also replaced the main data file.
Text modifications throughout. Replaced the data file according to the new release (https://github.com/zackbatist/open-archaeo/releases/tag/2024-02-03). Revised the abstract. Verified and corrected the bibliographic references.
Minor edits to readability and text flow

The interaction between project longevity, activity, and number of contributors is multifaceted [@fig-lifespan-rate].
Highly active projects (one commit per day or more) tend to be either very long-lived or very short-lived; few fall in the centre of the distribution.
Short-lived projects tend to be characterised by a 'spree' of activity (a high commit rate), while long-lived projects have a broader range of activity profiles.
The most "successful" projects according to open source norms (i.e. long-lived and active) are with few exceptions those projects with the largest contributor base in our dataset.
However, the modal project in the centre of the distribution is more modest, lasting around three years, maintained as by an individual or a small group, with around three commits per month.

@@

The prototypical and idyllic collaborative open source project ingrained into the common consciousness hypothetically comprises a core group of developers (often a single maintainer) that regularly commit new code, a wider network of collaborators that contribute through forks and pull requests, plus an active user base that create and comment on issues, who have indicated their support for the project by starring its repository.
It is unclear whether archaeological software developers actually aim to operate following this model, or whether it is even suitable for supporting what open science aims to achieve.
However it is clear that only a small number of open-archaeo projects operate according to this model.
The majority of projects are in fact short-lived, with few contributors and a small number of commits.
Use of GitHub's collaboration features is also generally low (@fig-collab-features), although the data also shows a divergence between the uptake of features that facilitate direct code contributions (forks, issues, pull requests), which have markedly zero-skewed distributions, versus more indirect, social media-like features (comments, stars), which are moderately well-used.
We hypothesise that this shows a preference for passive/reactive rather than active/proactive engagement with others work – a point we will return to in the conclusion.

>>>>

The interaction between project longevity, activity, and number of contributors is multifaceted (@fig-lifespan-rate).
Highly active projects (one commit per day or more) tend to be either very long-lived or very short-lived; few fall in the centre of the distribution.
Short-lived projects tend to be characterised by a 'spree' of activity (a high commit rate), while long-lived projects have a broader range of activity profiles.
The most "successful" projects according to open source norms (i.e. long-lived and active) are with few exceptions those projects with the largest contributor base in our dataset.
However, the modal project in the centre of the distribution is more modest, lasting around three years, maintained as by an individual or a small group, with around three commits per month.

@@

The prototypical open source project comprises a core group of developers (often a single maintainer) that regularly commit new code, a wider network of collaborators that contribute through forks and pull requests, plus an active user base that create and comment on issues, who have indicated their support for the project by starring its repository.
It is unclear whether archaeological software developers actually aim to operate following this model, or whether it is even suitable for supporting what open science aims to achieve.
However it is clear that only a small number of open-archaeo projects operate according to this model.
The majority of projects are in fact short-lived, with few contributors and a small number of commits.
Use of GitHub's collaboration features is also generally low (@fig-collab-features), although the data also shows a divergence between the uptake of features that facilitate direct code contributions (forks, issues, pull requests), which have markedly zero-skewed distributions, versus more indirect, social media-like features (comments, stars), which are moderately well-used.
We hypothesise that this shows a preference for passive/reactive rather than active/proactive engagement with others work – a point we will return to in the conclusion.

### 2024-02-06 analysis/_06-network.qmd
Minor edits to readability and text flow

<!-- Clustering within the network -->
To explore the topography of this network, we applied the edge–betweenness community detection method [@girvan2002] to delimit `r nrow(filter(clusters_n_repos, n > 1))` distinct clusters in the repository–repository network.
While many of these clusters are interconnected, some discrete components containing between 2-20 repositories appear as distinct from a primary core.
The core cluster is characterized by repositories whose contributors commit to projects other than their own, and it includes a smorgasbord of projects whose contributors share varied interests.

Likewise, we applied the same edge-betweenness community detection method to identify clusters in the user-user network, which reveal collaborative groupings.
We again see a complementary primary core connected to several more peripheral clusters, which are internally-cohesive and exhibit few connections with other peripheral clusters.
The central core bridges all the peripheral clusters.
The central core is not uniform, and comprises several relatively discrete sub-clusters representing collaborative sub-communities.
While these sub-clusters are internally cohesive, they exhibit enough connections to other members of the central core so as to not be considered as separate or peripheral clusters.

In both the repository-repository and user-user networks, the peripheral clusters correspond with either the connections surrounding specific projects or the series of repositories created by single individuals and sometimes also their close colleagues.
On the other hand, the central cores exhibit greater internal variety that may correspond with social connections and the formation of a complex software development community.
This is evident through the fact that many of the connections represented in the cores emerge from underlying professional partnerships, namely research labs (e.g. ISAA-Kiel) and special interest groups (CAA-SSLA).
Peripheral clusters that are connected to the central core by only a few relationships represent the sole (or perhaps initial) integration of lone developers into a broader community.

@@

Despite being a minority language, Blender packages are more central than all other package platforms on average, but this is a statistical anomaly caused by uneven sampling.
R is naturally the platform with the next highest average centrality since it serves as a _lingua franca_ that draws developers from across the discipline.
Many of the QGIS plugins add various specialized features to the extensible GIS platform, and are therefore developed by interdisciplinary teams, which explains its high rank.
Python projects, which tend to be infrastructural or are of interest to members of other fields, are also highly ranked in terms of average centrality.
It should be noted that while Blender and QGIS plugins are written using the Python language, our intent while categorizing platforms was to get a sense of the developer ecosystems in which archaeological software engineers participate, rather than to simply gauge the popularity of different languages [@batist2023 2].
Our findings support the notion that there are significant development patterns when working across different languages and platforms, and further analysis to qualify these observation is warranted.

>>>>

<!-- Clustering within the network -->
We delimited `r nrow(filter(clusters_n_repos, n > 1))` distinct clusters that outline the topography of the repository–repository network.
While many of these clusters are interconnected, some discrete components containing between 2-20 repositories appear as distinct from a primary core.
The core cluster is characterized by repositories whose contributors commit to projects other than their own, and it includes a smorgasbord of projects whose contributors share varied interests.

Clustering also reveals distinct collaborative networks within the user-user graph.
We again see a complementary primary core connected to several more peripheral clusters, which are internally-cohesive and exhibit few connections with other peripheral clusters.
The central core bridges all the peripheral clusters.
The central core is not uniform, and comprises several relatively discrete sub-clusters representing collaborative sub-communities.
While these sub-clusters are internally cohesive, they exhibit enough connections to other members of the central core so as to not be considered as separate or peripheral clusters.

In both the repository-repository and user-user networks, the peripheral clusters correspond with either the connections surrounding specific projects or the series of repositories created by single individuals and sometimes also their close colleagues.
On the other hand, the central cores exhibit greater internal variety that may correspond with social connections and the formation of a complex software development community.
This is evident through the fact that many of the connections represented in the cores emerge from more conventional professional networks, e.g. [ISAAKiel](https://isaakiel.github.io/), a working group centered around the University of Kiel), or [CAA-SSLA](https://sslarch.github.io/), a special interest group of the international scholarly society 'Computational and Quantitative Applications in Archaeology' (CAA) focused on scientific programming.
Peripheral clusters that are connected to the central core by only a few relationships represent the sole (or perhaps initial) integration of lone developers into a broader community.

@@

Despite being a minority language, Blender packages are more central than all other package platforms on average, but this is a statistical anomaly caused by uneven sampling.
R is naturally the platform with the next highest average centrality since it serves as a _lingua franca_ that draws developers from across the discipline.
Many of the QGIS plugins add various specialized features to the extensible GIS platform, and are therefore developed by interdisciplinary teams, which explains its high rank.
Python projects, which tend to be infrastructural or are of interest to members of other fields, are also highly ranked in terms of average centrality.
^[While Blender and QGIS plugins are written using the Python language, our intent while categorizing platforms was to get a sense of the developer ecosystems in which archaeological software engineers participate, rather than to simply gauge the popularity of different languages [@batist2023 2].]
Our findings support the notion that there are significant development patterns when working across different languages and platforms, and further analysis to qualify these observation is warranted.

### 2024-02-06 analysis/_07-conclusion.qmd
Minor edits to readability and text flow

Looking at the distribution of tags, we note that, among items listed as software, repositories tend to be focused on 
various means of identifying distribution patterns (spatial, temporal, statistical), 
calibrating data obtained from various instrumental methods (XRF, luminescence dating), 
supporting specialized finds analysis (zooarchaeology, palaeobotany, archaeogenetics), 
and supporting the collection and processing of archaeological materials.
These foci signify gaps in the archaeological toolbox that archaeologists recognized, and have attempted to fill, on their own terms.
In future studies, it would be interesting to examine how the purposeful expansion of open source tools corresponds with broader methodological trends apparent in publishing patterns or through other expressions of archaeological interest (e.g. social media posts, conference sessions).

@@

>>>>

Turning to the thematic distribution of open archaeology projects, we note that, among items listed as software, repositories tend to be focused on 
various means of identifying distribution patterns (spatial, temporal, statistical), 
calibrating data obtained from various instrumental methods (XRF, luminescence dating), 
supporting specialized finds analysis (zooarchaeology, palaeobotany, archaeogenetics), 
and supporting the collection and processing of archaeological materials.
These foci signify gaps in the archaeological toolbox that archaeologists recognized, and have attempted to fill, on their own terms.
In future studies, it would be interesting to examine how the purposeful expansion of open source tools corresponds with broader methodological trends apparent in publishing patterns or through other expressions of archaeological interest (e.g. social media posts, conference sessions).

@@

Overall, we found that the vast majority of projects lack any collective effort and activity tends to abruptly end shortly after work is initiated. Projects that are maintained by multiple contributors tend to be backed by funded initiatives who employ locally and socially engaged individuals, or are warranted by participation in a scholarly community exhibiting genuine need for a particular kind of resource. Moreover, we found that the individuals who play critical roles in supporting the archaeological open source community are precariously employed workers. Contrary to popular claims about open source being inherently distributed, resilient, and open-ended, we found that, overall, archaeological open source is actually quite centralized, fragile, and based nearly exclusively on existing professional connections and endeavours.
